Variation generation and mean fitness increase: \\ Necessity is the mother of genetic invention

### 2017-11-16 ms.tex
learning model

% Model
\section*{Models}

@@

In the following section we present three examples of the model (Eq.~\ref{eq:model}) that apply to mutation and migration.

@@

@@

% see Altenberg 2012 Ecological Monographs for a condition-dependent migration model
We have given several examples of variation-generating processes for which this principle applies -- namely mutation and migration -- but our model may apply to other processes as well.
For example, the \emph{reduction principle} applies to ecological models of dispersal, and~\citet{Gueijman2013} have demonstrated that even in homogeneous environments, fitness-associated dispersal increases the population mean fitness and is favored by selection over uniform dispersal.
% TODO see cultural transmission model in Lee's thesis
Similarly, if the transmission fidelity of culturally-transmitted traits depends on the type or fitness of the transmitting individual, we expect that our results will hold.

>>>>

% Model
\section*{Models}\label{sec:models}

@@

In the following section we present four examples of the model (Eq.~\ref{eq:model}) that apply to mutation, migration, and learning.

@@

% Learning model
\paragraph*{Learning model.}\label{sec:learn_model}

In our final example, we consider a large population and an integer phenotype $k$ where $1 \le k \le n$.
Individuals are characterized by their initial and mature phenotypes~\citep[pg.~94]{Boyd1985}.
Fitness is determined by the difference between the individual mature phenotype and the optimal phenotype:
denoting the optimal phenotype as $h$, the fitness of an individual with mature phenotype $k$ is $w_{|k-h|}$.

The initial phenotype is acquired by learning the mature phenotype of one of the parents (i.e. uni-parental vertical transmission).
In individuals with initial phenotype $k$, the mature phenotype is the same as the initial phenotype with probability $1-C_k$, and is modified by individual learning or \emph{exploration}~\citep{Borenstein2008} with probability $C_k$.
Such individual exploratory learning, which can be considered intentional or the result of transmission infidelity, modifies initial phenotype $k$ to mature phenotype $j$ with probability $M_{k,j}$.

For example, the case of \emph{symmetric individual learning}~\citep{Borenstein2008} can be represented by
\begin{equation}
M_{k,j} = \begin{cases}
\frac{1}{2b+1}, & \abs{k-j} \le b \\
0, & otherwise
\end{cases},
\end{equation}
where the breadth of exploration is $b \ge 0$, and phenotypes $0$ and $n$ "absorb" transitions to $k<0$ and $k>n$, respectively. 
For instance, for $n=5$ and $b=1$ we have
\begin{equation}
M = \begin{bmatrix}
2/3 & 1/3 & 0 & 0 & 0 \\
1/3 & 1/3 & 1/3 & 0 & 0 \\
0 & 1/3 & 1/3 & 1/3 & 0 \\
0 & 0 & 1/3 & 1/3 & 1/3 \\
0 & 0 & 0 & 1/3 & 2/3
\end{bmatrix}.
\end{equation}

Therefore, if the frequency of individuals with mature phenotype $k$ in the current generation is $f_k$, then the frequency in the next generation $f'_k$ is
\begin{equation}
f'_k = (1-C_k) f_k \frac{w_{\abs{k-h}}}{\bar{w}} \sum_{j=1}^{n}{C_j M_{k,j} \frac{w_{\abs{j-h}}}{\bar{w}} f_j },
\end{equation}
where $\bar{w} = \sum_{k=1}^{n}{f_k w_{\abs{k-h}}}$ is the population mean fitness.

@@

% see Altenberg 2012 Ecological Monographs for a condition-dependent migration model
We have given several examples of variation-generating processes for which this principle applies -- namely mutation, migration, and learning, see ~\nameref{sec:models} -- but our model may apply to other processes as well.
For example, the \emph{reduction principle} applies to ecological models of dispersal, and~\citet{Gueijman2013} have demonstrated that even in homogeneous environments, fitness-associated dispersal increases the mean fitness of diploid populations and is favored by selection over uniform dispersal.
Similarly, if the transmission fidelity of culturally-transmitted traits depends on the type or fitness of the transmitting individual, we expect that our results will hold (see~\nameref{sec:learn_model}).
